### 2015-07-07 narrative/data-reduction.tex

The classical workflow whereby the entire dataset is written to storage for
later analysis is no longer viable owing to the sheer amount of data being
generated. In future, it will be vital to take advantage of a priori
information when writing and reading from the SSIO layer in order to gain
performance and predictability, and to prioritize I/O so that the data that
is most useful for end-user can be written in the time available.

@@

In this section we describe some common methods to reduce and re-organize data.
There are many possibilities and the best choice will generally vary from case
to case but, once settled, will not change from run to run. It will be
important to understand when the time and resources required to identify and
perform the ``best methods'' outweighs any eventual gains. Another critical
research question concerns the quantification and control of the loss of
information resulting from refactoring the data and using a reduced 
dataset.

@@

Ideally, the scientists would perform the entire analysis stage {\em in situ} to
extract the relevant information and, in the unlikely event of this being
possible, would effectively circumvent the large data issue completely.  The
problem of course is that this is unlikely to be possible since, by their
nature, large scale simulations aim to discover new information that is often
hidden in the form of higher order effects amongst the data deluge. In
particular, this means if data thinning or truncation is applied willy nilly
then one is in danger of effectively throwing out the baby with the bath water.
One cannot store the entire data set in an easily accessible source, due to
sheer size, yet one cannot reduce it prior to archival without risking losing
the information one is trying to extract.  Viewed in this way, the problem
would appear intractable. However, the above discussion fails to recognize that
much of the data is redundant in the information theoretic sense.  That is to
say, the amount of information contained in the data is often significantly
less than the amount of data.  The difficulty stems from the fact that one does
not know about this redundancy in advance \emph{without the benefit of 
priori knowledge}. This knowledge can often be provided by the user or can 
be acquired through experience of dealing with differing runs of the same
code.

Application knowledge means one can sometimes achieve drastically superior
reduction in the total data compared with what one might achieve otherwise.
However, in the absence of such knowledge, we must equip the SSIO layers with
generic data reduction techniques and re-organization techniques.
Nevertheless, certain basic information on the semantics of the data is needed
and must be supported by the overall infrastructure.  Likewise, another
important consideration is knowledge regarding typical access patterns for data
in a given application. Scientific applications often read and analyze data in
a structured fashion and failure to respect this pattern, or to exploit it
through prefetching, can significantly increase the overall execution time of
the application. Accurate prefetching can only be achieved with access pattern
analysis. This requires an SSIO layer which is able to prefetch data in an
asynchronous fashion to take the bytes on the file.

@@

Research in the construction of pre-processing techniques is still in its
infancy, especially in context of exascale simulations. Techniques that have
been advocated range from simple difference filtering, bit-sorting, prefix
transformation through to simply reordering the data.  What these, and other
catch-all methods, fail to exploit is a priori information on the nature and
provenance of the data. Moreover, it is now widely recognized that any
opportunities for {\em in situ} analysis should be exploited as far as
possible. What is less well understood, is how one should incorporate a priori
information and how one can identify possibilities for {\em in situ} analysis.

We shall exploit a priori information in the preprocessing stage through the
use of relatively computationally intensive preprocessors that attempt to
extract structure in the data {\em in situ}. This approach takes advantage of
the increasingly wide cost gap between CPU operations versus storage-based
operations. The preprocessor would generally be employed in situ and is used in
a similar fashion to pre-processors in standard compression approaches such as
FPC \cite{BurtscherFPC} in which an XOR operation against the previous data is
used as a pre-processor.

Our previous
work~\cite{lakshminarasimhan2011compressing,lakshminarasimhan2011compressing,%
gong2012multi,jenkins2012byte,gong2013parlo,boyuka2014transparent,%
tang2014improving}
has focused on many of these well known techniques to not only reduce the data
but also re-organize the data. Previously we have examined:

@@

\paragraph{Proposed research approach:} We will be looking at a multi-throng
approach to data-refactoring. One approach allows the user to specify
information about their data to indicate if the data is spatially varying,
space-and time varying, just time varying. This allows us to store information
in NVRAM on each node (when and where necessary) to later preform compression
in both space and time.

We will examine the different techniques for
compression and re-organization along with examine techniques to use an
error-estimator (gradients) to create a multi-resolution view of the data which
can then be sampled, and placed in different quantiles of importance. These
will then be mapped over into the different layers of the storage hierarchy.

We will also investigate technique which can use ``plug-ins'' for detection of
features and store higher resolution data around the feature and lower
resolution values which are further away from the features. This
re-organization must occur without increasing the total amount of storage than
what the users intended.

@@

We will measure these methods using the fusion XGC simulation and the
corresponding analysis routines which they use for their largest simulations.
We will also work with Jeroen Tromp's SPECFM3D code, and the fusion GTC code
from Z. Lin. In all cases we will measure their impact from their current
simulations, and their projected simulations and analytics on exascale
machines. Since all three of these codes are part of the OLCF CAAR~\cite{CAAR}
project for early access, there is strong belief that the simulations will be
close to exascale simulations and we will be able to run these on prototypes of
our SSIO layer to examine the full impact of data-refactoring.

\paragraph{High Impact Challenges:} Since the greatest form of data reduction
on the SSIO layer would be to only write the code, and the input data, and then
recalculate all of the results, we want to understand how to write a minimal
set of information, and then place an auditor code in a docker container so
that data can be re-generated with an approximation of the calculation, but
only valid over a limited amount of time.This new approach to tackle the data
overload by using preprocessors in which information is not simply stored in
the form of raw data, but rather in the form of appropriate parameterized
algorithms. A priori information on the nature of the data is used to choose
the basic algorithm while the actual values of the data are used to tune the
parameters used in the reconstruction procedures. We shall refer to this
paradigm as an Auditor.

We will examine the use of several auditors for the simulations which we will
work with. One approach is to use a fluid-code to audit a kinetic code, which
can be a good approximation over short periods of time. Once there is a large
deviation (specified by the user) from the auditor to the original calculation,
the data will be saved. This is a local operation, since the deviations will
grow differently in space and in time.

>>>>

The classical workflow whereby the entire dataset is written to storage for
later analysis is no longer viable owing to the sheer amount of data being
generated. In future, it will be vital to take advantage of a priori
information when a) writing and reading from the SSIO layer in order to gain
performance and predictability, and to prioritize I/O so that the data that
is most useful for end-user can be written in the time available, and b) in
performing {\em in situ} analysis before storing the information.

@@

There are many possibilities for refactoring and the best choice will generally
vary from case to case but, once settled, will not change from run to run. It
will be important to understand when the time and resources required to
identify and perform the ``best methods'' outweighs any eventual gains. Another
critical research question concerns the quantification and control of the loss
of information resulting from refactoring the data and using a reduced dataset.

@@

Ideally, the scientists would perform the entire analysis stage {\em in situ}
to extract the relevant information and in so doing would effectively
circumvent the large data issue completely.  The catch, of course, is that this
is unlikely to be possible since, by their nature, large scale simulations aim
to discover new information that is often hidden in the form of higher order
effects amongst the data deluge. In particular, this means if data thinning or
truncation is applied willy nilly then one is in danger of effectively throwing
out the information one is seeking. It would seem that one cannot store the
entire data set in an easily accessible source, due to sheer size, yet one
cannot reduce it prior to archival without risking losing the information one
is trying to extract.  Viewed in this way, the problem would appear
intractable. However, the above discussion fails to recognize that much of the
data is redundant in an information theoretic sense.  That is to say, the
amount of information contained in the dataset is often significantly less than
the amount of data.  The difficulty stems from the fact that one does not know
about this redundancy in advance \emph{without the benefit of priori
knowledge}. This knowledge can often be provided by the user or can be acquired
through experience of dealing with differing runs of the same code.

Deep application knowledge means one can sometimes achieve drastically superior
reduction in the total data compared with what one might achieve otherwise.
However, even in the absence of such high level knowledge, we must equip the
SSIO layers with generic data reduction techniques and re-organization
techniques. For instance, certain basic information on the semantics of the
data is needed and must be supported by the overall infrastructure. Likewise,
information regarding typical access patterns for data in a given application
should be placed in the SSIO layers. Scientific applications often read and
analyze data in a structured fashion and failure to respect this pattern, or to
exploit it through prefetching, can significantly increase the overall
execution time of the application. Accurate prefetching can only be achieved
with access pattern analysis. This requires an SSIO layer which is capable of
prefetching data in an asynchronous fashion to take the bytes on the file.

@@

Research in the construction of pre-processing techniques is still in its
infancy, especially in context of exascale simulations. Techniques that have
been advocated range from simple difference filtering, bit-sorting, prefix
transformation through to simply reordering the data.  What these, and other
catch-all methods, fail to exploit is a priori information on the nature and
provenance of the data.

Our previous
work~\cite{lakshminarasimhan2011compressing,lakshminarasimhan2011compressing,%
gong2012multi,jenkins2012byte,gong2013parlo,boyuka2014transparent,%
tang2014improving}
has focused on several alternative techniques to not only reduce the data
but also re-organize the data. Previously we have examined:

@@

\paragraph{Proposed research approach:} It is now widely recognized that any
opportunities for {\em in situ} analysis should be exploited as far as
possible. What is less well understood, is how one should incorporate a priori
information and how one can identify possibilities for {\em in situ} analysis.
A preprocessor would be employed in situ and used in a similar fashion to
pre-processors in standard compression approaches such as FPC
\cite{BurtscherFPC} in which an XOR operation against the previous data is used
as a pre-processor.  We shall exploit a priori information in the preprocessing
stage through the use of relatively computationally intensive preprocessors
(compared with FPC) that seek to extract structure in the data {\em in situ}.
This approach takes advantage of the increasingly wide cost gap between CPU
operations versus storage-based operations.

Ideally, the scientists would perform the entire analysis stage {\em in situ}
to extract the relevant information and in so doing would effectively
circumvent the large data issue completely.  The problem of course is that this
is unlikely to be possible since, by their nature, large scale simulations aim
to discover new information that is often hidden in the form of higher order
effects amongst the data deluge. In particular, this means if data thinning or
truncation is applied willy nilly then one is in danger of effectively throwing
out the information one is seeking. It would seem that one cannot store the
entire data set in an easily accessible source, due to sheer size, yet one
cannot reduce it prior to archival without risking losing the information one
is trying to extract.  Viewed in this way, the problem would appear
intractable. However, the above discussion fails to recognize that much of the
data is redundant in an information theoretic sense.  That is to say, the
amount of information contained in the dataset is often significantly less than
the amount of data.  The difficulty stems from the fact that one does not know
about this redundancy in advance \emph{without the benefit of priori
knowledge}. This knowledge can sometimes be provided by the user, can be
acquired through experience of dealing with differing runs of the same code, or
acquired during a run by adaptively selecting and tuning the algorithm based on
metrics during runtime.

We will assemble a suite of general techniques for refactoring of data that
will be embedded in the system and can be selected either by the user or by the
application itself. The techniques will range from crude catch all methods such
as those employed by general compression routines such as FPC, through to
multi-resolution analysis using wavelets, including the methods on which we
have worked (cited above).  These will then be mapped over into the different
layers of the storage hierarchy.  We will also cater for user supplied
``plug-ins'' for detection of features and store higher resolution data around
the feature and lower resolution values which are further away from the
features. The plug-ins could be used to override the general embedded
techniques and would provide a mechanism for the transfer of a priori knowledge
from the user to the hierarchy without imposing a burden on either.

@@

We will measure these methods using the fusion XGC simulation and the
corresponding analysis routines which they use for their largest simulations.
We will also work with Jeroen Tromp's SPECFM3D~\cite{SPECFEM3D} code, and the
fusion GTC~\cite{klasky2003grid} code.  In all cases we will measure their
impact from their current simulations, and their projected simulations and
analytics on exascale machines. Since all three of these codes are part of the
OLCF CAAR~\cite{CAAR} project for early access, there is strong belief that the
simulations will be close to exascale simulations and we will be able to run
these on prototypes of our SSIO layer to examine the full impact of
data-refactoring.

\paragraph{High Impact Challenges:} Maximal data reduction on the SSIO layer
would result from storing only the code and the input parameters, and simply
recalculating all of the results. This extreme is not feasible in practice
owingto the size of the computation involved, but sheds light on the path to
achieving a step change in data reduction. We shall investigate the possibility
of writing a minimal set of information, and placing an ``auditor'' code in a
docker container so that data can be re-generated with a {\em local}
approximation of the full simulation. This is an entirely new approach to
tackle the data overload by using preprocessors in which information is not
simply stored in the form of raw data, but rather in the form of appropriate
parameterized algorithms. A priori information on the nature of the data is
used to choose the basic algorithm while the actual values of the data are used
to tune the parameters used in the reconstruction procedures. We shall refer to
this paradigm as an Auditor.  We will examine the use of several auditors for
the simulations which we will work with. One approach is to use a fluid-code to
audit a kinetic code, which can be a good approximation over short periods of
time. Once there is a large deviation (specified by the user) from the auditor
to the original calculation, the data will be saved. This is a local operation,
since the deviations will grow differently in space and in time.

%%%%%%%%%%%%%
We will be looking at a multi-throng
approach to data-refactoring. One approach allows the user to specify
information about their data to indicate if the data is spatially varying,
space-and time varying, just time varying. This allows us to store information
in NVRAM on each node (when and where necessary) to later preform compression
in both space and time.
